|

### 2021-12-14 01-Chp-Background.Rmd
Range of edits incl: - Better figure captions - Better description of proportional biases - Julian comments to Discussion
End of day:

@@

In terms of the central research of this thesis, RCTs of statin therapy can be used to provide indirect evidence for the effect of reducing blood LDL-c levels on dementia risk. However, RCTs may be infeasible if the outcome of interest is one with a long prodomal period, such as dementia (see Section \@ref(underlying-pathologies)), as they would require extremely long and costly follow-up.[@ritchie2015] It is no surprise then that the two previous trials providing evidence on the effect of statins on dementia risk, identified by a recent Cochrane review,[@mcguinness2016] are in fact trials of statins for the prevention of coronary related outcomes.

>>>>

<!-- TODO Need some details on risk factors here - and consider spacing differentl -->

@@

In terms of the central research of this thesis, RCTs of statin therapy can be used to provide indirect evidence for the effect of reducing blood LDL-c levels on dementia risk. However, RCTs may be infeasible if the outcome of interest is one with a long prodromal period, such as dementia (see Section \@ref(underlying-pathologies)), as they would require extremely long and costly follow-up.[@ritchie2015] It is no surprise then that the two previous trials providing evidence on the effect of statins on dementia risk, identified by a recent Cochrane review,[@mcguinness2016] are in fact trials of statins for the prevention of coronary related outcomes.

### 2021-12-14 06-Chp-IPD-Analysis.Rmd
End of day:

Previous work has suggested a difference in the effect of lipids on dementia risk based on participant age and sex,[@ancelin2013; @mielke2010] though the systematic review presented in Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading) did not identify strong evidence for an effect of sex or age at lipid measurement on the relationship between lipids and dementia. However, the number of included studies in the lipid fraction meta-analyses was small, due to both the relatively small number of studies reporting on the lipids exposure and the poor reporting of summary statistics of participant characteristics in those that did. Additionally, best practice guidance recommends against basing the decision to perform an IPD meta-analysis on the between-study heterogeneity in a summary data meta-analysis, as similar distributions of participant covariates across studies may mask a true effect.[@riley2020]

@@

Missing variables were originally intended to be addressed using a previously published method.[@fibrinogenstudiescollaboration2009] Here, the correlation between the fully-adjusted and partially-adjusted estimate in cohorts with the full set of covariates is used to estimate the fully-adjusted effect in those cohorts missing covariates. However, this method requires several large cohorts to contain the full set of covariates, a condition this analysis failed to meet given the very low response rate. As such, two other common approaches were employed.[@fibrinogenstudiescollaboration2009] In the primary analysis, all cohorts were adjusted for the set of common covariates across cohorts (Model 1: age, sex, smoking, alcohol, diabetes). As a sensitivity analysis, cohorts with the full complement of covariates (Whitehall II and EPIC) were analysed using a maximally-adjusted model (Model 2: Model 1, further adjusted for education, ethnicity, prevalent ischemic heart disease (IHD) and BMI). Results between the common-set-adjusted (Model 1) and maximally-adjusted (Model 2) models were then compared.

@@

@@

The Caerphilly Prospective Study (CaPS) is a longitudinal study of men in South Wales, UK.[@zotero-15398;@elwood2013] Blood lipids (TC, LDL-c, HDL-c and TG) were measured at baseline in 1979-1983, and from Phase III (1989-1993) onwards, a battery of cognitive tests were introduced. A dementia was derived using data obtained during Phase V (2002-2004), giving between 19-25 years of follow-up. The cohort contains data on dementia outcomes sub-classified as vascular and non-vascular dementia. Data was available for all covariates of interest except for education, BMI and _Apo_$\mathcal{E}4$ status.

@@

Estimates from the common-set-adjusted model (Model 1: adjusted for age, sex, smoking, alcohol, and diabetes) were comparable to the fully-adjusted model (Model 2: Model 1 further adjusted for ethnicity, prevalent ischemic heart disease, education and BMI) for the effect of lipids on all-cause dementia in cohorts reporting all covariates of interest (Figure \@ref(fig:ipdModelComparison)).

@@

<!----------------------------------------------------------------------------->
(ref:interactionDementiaAge-cap)  __Meta-analysis of age-lipid interaction terms for all-cause dementia__ - Age was grouped into 5-year age bands, and the effect estimatse presented represent the OR per 1-step increase in age group. Estimates were obtained using the common-set-adjusted model (Model 1: age, sex, smoking, alcohol, diabetes).

@@

<!----------------------------------------------------------------------------->
(ref:interactionDementiaSex-cap) __Meta-analysis of sex-lipid interaction terms for all-cause dementia__ Estimates were obtained using the common-set-adjusted model (Model 1: age, sex, smoking, alcohol, diabetes) and refer to the effect of male gender on the lipid/all-cause dementia association.

@@

<!----------------------------------------------------------------------------->
(ref:interactionVascularAge-cap) __Meta-analysis of age-lipid interaction terms for vascular dementia__ - Age was grouped into 5-year age bands, and the effect estimatse presented represent the OR per 1-step increase in age group. Estimates were obtained using the common-set-adjusted model (Model 1: age, sex, smoking, alcohol, diabetes).

>>>>

Previous work has suggested a difference in the effect of lipids on dementia risk based on participant age and sex.[@ancelin2013; @mielke2010] However, the systematic review presented in Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading) did not identify strong evidence for an effect of sex or age at lipid measurement on the relationship between lipids and dementia. The number of included studies in the lipid fraction meta-analyses was small, due to both the relatively small number of studies reporting on the lipids exposure and the poor reporting of summary statistics of participant characteristics in those that did. Additionally, best practice guidance recommends against basing the decision to perform an IPD meta-analysis on the between-study heterogeneity in a summary data meta-analysis, as similar distributions of participant covariates across studies may mask a true effect.[@riley2020]

@@

Missing variables were originally intended to be addressed using a previously published method.[@fibrinogenstudiescollaboration2009] Here, the correlation between the fully-adjusted and partially-adjusted estimate in cohorts with the full set of covariates is used to estimate the fully-adjusted effect in those cohorts missing covariates. However, this method requires several large cohorts to contain the full set of covariates, a condition this analysis failed to meet given the very low response rate. As such, two other common approaches were employed.[@fibrinogenstudiescollaboration2009] In the primary analysis, all cohorts were adjusted for the set of common covariates across cohorts (Model 1: age, sex, smoking, alcohol, education, diabetes). As a sensitivity analysis, cohorts with the full complement of covariates (Whitehall II and EPIC) were analysed using a maximally-adjusted model (Model 2: Model 1, further adjusted for ethnicity, prevalent ischemic heart disease and BMI). Results between the common-set-adjusted (Model 1) and maximally-adjusted (Model 2) models were then compared.

@@

<!-- REVIEW You need to justify why RE and FE for main and interaction terms -->

@@

The Caerphilly Prospective Study (CaPS) is a longitudinal study of men in South Wales, UK.[@zotero-15398;@elwood2013] Blood lipids (TC, LDL-c, HDL-c and TG) were measured at baseline in 1979-1983, and from Phase III (1989-1993) onwards, a battery of cognitive tests were introduced. A dementia was derived using data obtained during Phase V (2002-2004), giving between 19-25 years of follow-up. The cohort contains data on dementia outcomes sub-classified as vascular and non-vascular dementia. Data was available for all covariates of interest except for ethnicity, BMI, prevalent IHD and _Apo_$\mathcal{E}4$ status. While the CaPS study collected data on height/weight (from which BMI can be calculated as $\frac{weight}{height^2}$) and prevalent heart disease, these were not available from the DPUK version of the CaPS data.

@@

Estimates from the common-set-adjusted model (Model 1: adjusted for age, sex, smoking, alcohol, education and diabetes) were comparable to the fully-adjusted model (Model 2: Model 1 further adjusted for ethnicity, prevalent ischemic heart disease, and BMI) for the effect of lipids on all-cause dementia in cohorts reporting all covariates of interest (Figure \@ref(fig:ipdModelComparison)).

@@

<!----------------------------------------------------------------------------->
(ref:interactionDementiaAge-cap)  __Meta-analysis of age-lipid interaction terms for all-cause dementia__ - Age was grouped into 5-year age bands, and the effect estimatse presented represent the OR per 1-step increase in age group. Estimates were obtained using the common-set-adjusted model (Model 1: age, sex, smoking, alcohol, education, diabetes).

@@

<!----------------------------------------------------------------------------->
(ref:interactionDementiaSex-cap) __Meta-analysis of sex-lipid interaction terms for all-cause dementia__ Estimates were obtained using the common-set-adjusted model (Model 1: age, sex, smoking, alcohol, education, diabetes) and refer to the effect of male gender on the lipid/all-cause dementia association.

@@

<!----------------------------------------------------------------------------->
(ref:interactionVascularAge-cap) __Meta-analysis of age-lipid interaction terms for vascular dementia__ - Age was grouped into 5-year age bands, and the effect estimatse presented represent the OR per 1-step increase in age group. Estimates were obtained using the common-set-adjusted model (Model 1: age, sex, smoking, alcohol, education, diabetes).

### 2021-12-14 07-Chp-Triangulation.Rmd
End of day:

These causal questions were chosen for several reasons. Firstly, the work presented in the previous chapters had identified some evidence for an association in these exposure/outcome pairs. Secondly, given the long prodomal period between the onset of physiological changes in the brain and presentation of dementia symptoms, it seems likely that conditions during the mid-life period are particularly important. Finally, examining the causal impact of a lipid fraction on a specific outcome (e.g. Alzheimer's disease), rather than a composite outcome (e.g. all-cause dementia), focuses on an single causal pathway rather than including several conditions in the outcome definition that likely have very different mechanisms of disease. In addition, previous work presented in this thesis illustrated that the component conditions (in this case Alzheimer's disease and vascular dementia) are likely to be subject to very different confounding structures (see Section \@ref(cprd-confounding-by-ind)), and the use of a composite all-cause outcome in the triangulation framework may mask this.

@@

Effects were standardised to refer to the risk resulting from a "reduction" in the lipid fraction. For both non-randomised studies and Mendelian randomisation analysis of lipid levels, the association is usually reported per 1-SD increase in the lipid fraction. In this case, the effect estimates were inverted to ensure consistency across study designs.

@@

@@

Reasonable values for the position and spread of the prior distributions were informed by a previously-reported expert bias elicitation exercise.[@turner2009] Using open data from that study, I calculated the mean values for the mean and variance of the adjustment distributions across all biases considered by that exercise (see Appendix \@ref(appendix-adjustment-values) for more details on how these were calculated). For the purposes of this illustrative analysis and to highlight the generalised nature of the code used to perform the analysis (see Section \@ref(tri-software)), I used the calculated values to inform two scenarios of modifying distributions (Table \@ref(tab:priorsAdd-table)) for additive biases, and a single set for proportional biases. In the first scenario, the relationship between the position assigned to each additive judgement is linear. In the second scenario, the step from "Moderate" to "High" is twice that from "Low" to "Moderate". Across both additive and proportional biases, the variance of the "High" judgement distribution was defined to be greater than that of the "Moderate" judgement.

@@

####	Combining in a bias-adjusted meta-analysis

@@

&nbsp;<!----------------------------------------------------------------------->  
(ref:biasDirectionSingle-cap) Bias direction plot for a single study showing the predicted direction of bias in each domain. Additive biases are characterised using arrows, while the proportional biases are denoted using $<$ and $>$. A question mark indicates that the direction of bias was unpredictable.

@@

&nbsp;<!----------------------------------------------------------------------->  
(ref:singleIndirect-caption) Assessment of indirectness for a single example study.

(ref:singleIndirect-scaption) singleIndirect

@@

The unadjusted estimate was `r single_unadjusted `. After accounting for the biases and indirectness described, the adjusted result was `r single_adjusted `.

@@

(ref:ldlAdBiasDirection-cap) Bias direction plot, summarising the internal biases related to effect of LDL-c on Alzheimer's disease, stratified by study design.

@@

(ref:fpLdlAd-cap) Random effects meta-analysis of the association of midlife LDL-c with Alzheimer's disease, using unadjusted and bias-/indirectness-adjusted results.

(ref:fpLdlAd-scap) Random effects meta-analysis using unadjusted and bias-/indirectness-adjusted results.

@@

<!----------------------------------------------------------------------------->
(ref:tgVadBiasDirection-cap) Bias direction plot, summarising the internal biases related to effect of triglycerides on vascular dementia, stratified by study design.

@@

<!----------------------------------------------------------------------------->
(ref:fpTgVad-cap) Random effects meta-analysis of the association of midlife triglycerides and vascular dementia, using unadjusted and bias-/indirectness-adjusted (Scenario 1) results.

(ref:fpTgVad-scap) shortcap

@@

@@

>>>>

These causal questions were chosen for several reasons. Firstly, the work presented in the previous chapters had identified some evidence for an association in these exposure/outcome pairs. Secondly, given the long prodromal period between the onset of physiological changes in the brain and presentation of dementia symptoms, it seems likely that conditions during the mid-life period are particularly important. Finally, examining the causal impact of a lipid fraction on a specific outcome (e.g. Alzheimer's disease), rather than a composite outcome (e.g. all-cause dementia), focuses on an single causal pathway rather than including several conditions in the outcome definition that likely have very different mechanisms of disease. In addition, previous work presented in this thesis illustrated that the component conditions (in this case Alzheimer's disease and vascular dementia) are likely to be subject to very different confounding structures (see Section \@ref(cprd-confounding-by-ind)), and the use of a composite all-cause outcome in the triangulation framework may mask this.

@@

Effects were standardised to refer to the risk resulting from a "reduction" in the lipid fraction. For both non-randomised studies and Mendelian randomisation analysis of lipid levels, the association is usually reported in terms of an increase in the lipid fraction. In this case, the effect estimates were inverted to ensure consistency across study designs.

@@

<!-- REVIEW How did you deal with the statin studies? Did you adjust for the average reduction in LDL-c for average dose? Need to add clarification here.

This should be an issue, as different effect sof statins will be adjusted for bia indirectness aspect of framework.

Move to questions
-->

@@

Reasonable values for the position and spread of the prior distributions were informed by a previously reported expert bias elicitation exercise.[@turner2009] Using open data from that study, I calculated the mean values for the mean and variance of the adjustment distributions across all biases considered by that exercise (see Appendix \@ref(appendix-adjustment-values) for more details on how these were calculated). For the purposes of this illustrative analysis and to highlight the generalised nature of the code used to perform the analysis (see Section \@ref(tri-software)), I used the calculated values to inform two scenarios of modifying distributions (Table \@ref(tab:priorsAdd-table)) for additive biases, and a single set for proportional biases. In the first scenario, the relationship between the position assigned to each additive judgement is linear. In the second scenario, the step from "Moderate" to "High" is twice that from "Low" to "Moderate". Across both additive and proportional biases, the variance of the "High" judgement distribution was defined to be greater than that of the "Moderate" judgement.

@@

####	Combining in a bias-/indirectness-adjusted meta-analysis

@@

<!-- REVIEW YBS - I may have got confused but when you say favour the intervention I am assuming you mean a “protective effect” e.g. OR <1. But if there is differential misclassification so that statin use increases diagnosis of vascular dementia that would show a deleterious effect under the null hence shift to right e.g. OR >1. Perhaps I have misunderstood favour as < risk when in fact you mean > risk. Can you clarify?

This is true for the VaD outcome, but in this case we are interested in the Alzheimer's diease outcome, which would result in under-reporting of AD when on statins and therefore favour the intervention.-->

&nbsp;<!----------------------------------------------------------------------->  
(ref:biasDirectionSingle-cap) __Bias direction plot for a single study__ - Summary of result and the predicted direction of bias in each domain of bias for the effect of statin use on Alzheimer's dementia, as estimated in Chapter \@ref(cprd-analysis-heading). Additive biases are characterised using arrows, while the proportional biases are denoted using $<$ and $>$. A question mark indicates that the direction of bias was unpredictable.

@@

<!-- TODO Email Jonathan about how ROBINS-I deals with reverse triangulation? -->

&nbsp;<!----------------------------------------------------------------------->  
(ref:singleIndirect-caption) __Assessment of indirectness for a single example study__

(ref:singleIndirect-scaption) Assessment of indirectness for a single example study

@@

The unadjusted estimate was `r single_unadjusted `. After accounting for the biases and indirectness described, the adjusted result was `r single_adjusted `, representing a minimal shift in the position of the effect estimate but a substantial decrease in precision.

@@

(ref:ldlAdBiasDirection-cap) __Bias direction plot for LDL-c/Alzheimer's disease__ - Summary of the unadjusted results and internal biases related to effect of LDL-c on Alzheimer's disease, stratified by study design.

@@

(ref:fpLdlAd-cap) __Bias-/indirectness-adjusted meta-analysis of midlife LDL-c on Alzheimer's disease__ - Random-effects meta-analysis of the association of midlife LDL-c with Alzheimer's disease, using unadjusted and bias-/indirectness-adjusted results. Note the substantial decrease in the weight assigned to the result of the CPRD analysis between the two plots, which is due to the high levels of bias/indirectness in this result as described in the above sections.

(ref:fpLdlAd-scap) Bias-/indirectness-adjusted meta-analysis of midlife LDL-c on Alzheimer's disease

@@

<!----------------------------------------------------------------------------->
(ref:tgVadBiasDirection-cap) __Bias direction plot for triglycerides/vascular dementia__ - Summary of the unadjusted results and internal biases related to effect of triglycerides on vascular dementia, stratified by study design.

@@

<!----------------------------------------------------------------------------->
(ref:fpTgVad-cap) __Bias-/indirectness-adjusted meta-analysis of midlife triglycerides on vascular dementia__ - Random effects meta-analysis of the association of midlife triglycerides and vascular dementia, using unadjusted and bias-/indirectness-adjusted (Scenario 1) results.

(ref:fpTgVad-scap) Bias-/indirectness-adjusted meta-analysis of midlife triglycerides on vascular dementia

@@

In the case of common outcomes, an extra step may be needed in the framework to assesses whether it is reasonable to synthesis across different effect estimates.

@@

<!-- REVIEW YBS I was struck by how much the 95% Cis were widened with CPRD. Is there a risk we are adding to much uncertainty and therefore down weighting this too much? -->
